---
title: "Vis: Data Visualization Basics"
author: Zach del Rosario
date: 2020-05-03
output: github_document
time: 10
reading: 30
---

# Vis: Data Visualization Basics

*Purpose*: The most powerful way for us to learn about a dataset is to
*visualize the data*. Throughout this class we will make extensive use
of the *grammar of graphics*, a powerful graphical programming *grammar*
that will allow us to create just about any graph you can imagine!

*Reading*: [Data Visualization
Basics](https://rstudio.cloud/learn/primers/1.1). *Note*: In RStudio use
`Ctrl + Click` (Mac `Command + Click`) to follow the link. *Topics*:
`Welcome`, `A code template`, `Aesthetic mappings`. *Reading Time*: \~
30 minutes

```{r setup, include=FALSE}
library(tidyverse)
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

### **q1** Inspect the `diamonds` dataset. What do the `cut`, `color`, and `clarity` variables mean?

*Hint*: We learned how to inspect a dataset in `e-data-00-basics`!

```{r q1-task, eval=FALSE}
?diamonds
```

-   cut = quality of cut
-   color = quality of color
-   clarity = how clear is the diamond

### **q2** Use your "standard checks" to determine what variables the dataset has.

Now that we have the list of variables in the dataset, we know what we
can visualize!

### **q3** Using `ggplot`, visualize `price` vs `carat` with points. What trend do you observe?

*Hint*: Usually the language `y` vs `x` refers to the `vertical axis` vs
`horizontal axis`. This is the opposite order from the way we often
specify `x, y` pairs. Language is hard!

```{r q3-task}
## TODO: Complete this code
ggplot(diamonds) + geom_point(mapping = aes(x = carat, y = price))
```

**Observations**:

-   There is a distinct positive correlation between price and carat.
-   The middle carat values seem to greater price variation, likely due
    to the other quality measures.

## A note on *aesthetics*

The function `aes()` is short for *aesthetics*. Aesthetics in ggplot are
the mapping of variables in a dataframe to visual elements in the graph.
For instance, in the plot above you assigned `carat` to the `x`
aesthetic, and `price` to the `y` aesthetic. But there are *many more*
aesthetics you can set, some of which vary based on the `geom_` you are
using to visualize. The next question will explore this idea more.

### **q4** Create a new graph to visualize `price`, `carat`, and `cut`simultaneously.

*Hint*: Remember that you can add additional aesthetic mappings in
`aes()`. Some options include `size`, `color`, and `shape`.

```{r q4-task}
## TODO: Complete this code
ggplot(diamonds) + 
  geom_point(mapping = aes(
    x = carat,
    y = price,
    color = cut
  ))
```

**Observations**:

-   I notice that this layout of the plot is not the best possible way
    to visualize this data, but it seems to work pretty well.
-   I notice that R automatically interpreted the data type of `cut`
    properly into a color scale.

<!-- include-exit-ticket -->

# Exit Ticket

<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit
ticket survey**, [linked
here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-vis00-basics-assignment.Rmd).
