---
title: "Vis: Bar Charts"
author: Zach del Rosario
date: 2020-05-13
output: github_document
time: 30
reading: 30
---

# Vis: Bar Charts

*Purpose*: *Bar charts* are a key tool for EDA. In this exercise, we'll
learn how to construct a variety of different bar charts, as well as
when---and when *not*---to use various charts.

*Reading*: [Bar Charts](https://rstudio.cloud/learn/primers/3.2)
*Topics*: (All topics) *Reading Time*: \~30 minutes

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)

```

### **q1** In the reading, you learned the relation between `geom_bar()` and `geom_col()`. Use that knowledge to convert the following `geom_bar()` plot into the same visual using `geom_col()`.

```{r q1-task}
transcount <- count(mpg, trans)
transcount %>%
  ggplot() +
  geom_col(mapping = aes(x = trans, y = n))
```

The reading mentioned that when using `geom_col()` our x-y data should
be `1-to-1`. This next exercise will probe what happens when our data
are not `1-to-1`, and yet we use a `geom_col()`. Note that a
[one-to-one](https://en.wikipedia.org/wiki/Injective_function) function
is one where each input leads to a single output. For the `mpg` dataset,
we can see that the pairs `cty, hwy` clearly don't have this one-to-one
property:

```{r mpg-not-1-to-1}
## NOTE: Run this chunk for an illustration
mpg %>% filter(cty == 20)
```

### **q2** The following code attempts to visualize `cty, hwy` from `mpg` using

`geom_col()`. There's something fishy about the `hwy` values; what's
wrong here?

*Hint*: Try changing the `position` parameter for `geom_col()`.

```{r q2-task}
mpg %>%
  ggplot(aes(x = cty, y = hwy)) +
  geom_col(position = "dodge")
```

**Observations**: - When "dodge" is enabled, I can actually see the
correlation between hwy and cty, rather than just a mess of overlapping
bars.

A more standard way to visualize this kind of data is a *scatterplot*,
which we'll study later. For now, here's an example of a more effective
way to visualize `cty` vs `hwy`:

```{r q2-compare}
## NOTE: Run this chunk for an illustration
mpg %>%
  ggplot(aes(cty, hwy)) +
  geom_point()
```

### **q3** The following are two different visualizations of the `mpg` dataset.

Document your observations between the `v1` and `v2` visuals. Then,
determine which---`v1` or `v2`---enabled you to make more observations.
What was the difference between the two visuals?

```{r q4-v1}
## TODO: Run this code without changing, describe your observations on the data
mpg %>%
  ggplot(aes(class, fill = class)) +
  geom_bar()
```

**Observations**: - The x axis and fill are describing the same
attribute, making the plot more complex than it needs to be, given what
it is presenting.

```{r q4-v2}
## TODO: Run this code without changing, describe your observations on the data
mpg %>%
  ggplot(aes(class, fill = drv)) +
  geom_bar()
```

**Observations**: - The plot is again using fill, but this time
presenting another useful attribute of the observations.

**Compare `v1` and `v2`**:

-   Which visualization---`v1` or `v2`---enabled you to make more
    observations?
    -   `v2` definitely allowed me to make more observations, not just
        about the number of different vehicle classes, but how that
        relates to their number of driven wheels.
-   What was the difference between `v1` and `v2`?
    -   `v1` used fill for the same attribute as the x axis, while `v2`
        took advantage of fill for a different attribute.

### **q4** The following code has a bug; it does not do what its author intended.

Identify and fix the bug. What does the resulting graph tell you about
the relation between `manufacturer` and `class` of cars in this dataset?

*Note*: I use a `theme()` call to rotate the x-axis labels. We'll learn
how to do this in a future exercise.

```{r q4-task}
mpg %>%
  ggplot(aes(x = manufacturer, fill = class)) +
  geom_bar(position = "dodge") +
  theme(axis.text.x = element_text(angle = 270, vjust = 0.5, hjust = 0))
```

**Observations** - The plot shows me how each automaker's lineup is
split between different vehicle classes. The error was that the
`position` call was inside `ggplot()` instead of `geom_bar`.

### **q5** The following graph is hard to read. What other form of faceting would make the visual more convenient to read? Modify the code below with your suggested improvement.

```{r q5-task}
mpg %>%
  ggplot(aes(x = cyl)) +
  geom_bar() +
  facet_wrap(~ manufacturer, )
```

<!-- include-exit-ticket -->

# Exit Ticket

<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit
ticket survey**, [linked
here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-vis01-bar-charts-assignment.Rmd).
