---
title: "Data: Deriving Quantities"
author: Zachary del Rosario
date: 2020-05-07
output: github_document
time: 15
reading: 60
---

# Data: Deriving Quantities

*Purpose*: Often our data will not tell us *directly* what we want to
know; in these cases we need to *derive* new quantities from our data.
In this exercise, we'll work with `mutate()` to create new columns by
operating on existing variables, and use `group_by()` with `summarize()`
to compute aggregate statistics (summaries!) of our data.

*Reading*: [Derive Information with
dplyr](https://rstudio.cloud/learn/primers/2.3) *Topics*: (All topics,
except *Challenges*) *Reading Time*: \~60 minutes

*Note*: I'm considering splitting this exercise into two parts; I
welcome feedback on this idea.

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)

```

### **q1** What is the difference between these two versions? How are they the same? How are they different?

```{r q1-task}
## Version 1
filter(diamonds, cut == "Ideal")
## Version 2
diamonds %>% filter(cut == "Ideal")
```

The reading mentioned various kinds of *summary functions*, which are
summarized in the table below:

## Summary Functions

| Type     | Functions                                            |
|----------|------------------------------------------------------|
| Location | `mean(x), median(x), quantile(x, p), min(x), max(x)` |
| Spread   | `sd(x), var(x), IQR(x), mad(x)`                      |
| Position | `first(x), nth(x, n), last(x)`                       |
| Counts   | `n_distinct(x), n()`                                 |
| Logical  | `sum(!is.na(x)), mean(y == 0)`                       |

### **q2** Using `summarize()` and a logical summary function, determine the number of rows with `Ideal` `cut`. Save this value to a column called `n_ideal`.

```{r q2-task}
df_q2 <-
  diamonds %>%
    filter(cut == "Ideal") %>%
    summarize(n_ideal = n())
```

The following test will verify that your `df_q2` is correct:

```{r q2-tests}
## NOTE: No need to change this!
assertthat::assert_that(
  assertthat::are_equal(
    df_q2 %>% pull(n_ideal),
    21551
  )
)
print("Great job!")
```

### **q3** The function `group_by()` modifies how other dplyr verbs function. Uncomment the `group_by()` below, and describe how the result changes.

When grouping by color and clarity, the result includes the average
price for each color/clarity combination, instead of one overall price
average.

```{r q3-task}
diamonds %>%
  group_by(color, clarity) %>%
  summarize(price = mean(price))
```

### Vectorized Functions

| Type            | Functions                                                                            |
|-----------------------|------------------------------------------------|
| Arithmetic ops. | `+, -, *, /, ^`                                                                      |
| Modular arith.  | `%/%, %%`                                                                            |
| Logical comp.   | `<, <=, >, >=, !=, ==`                                                               |
| Logarithms      | `log(x), log2(x), log10(x)`                                                          |
| Offsets         | `lead(x), lag(x)`                                                                    |
| Cumulants       | `cumsum(x), cumprod(x), cummin(x), cummax(x), cummean(x)`                            |
| Ranking         | `min_rank(x), row_number(x), dense_rank(x), percent_rank(x), cume_dist(x), ntile(x)` |

### **q4** Comment on why the difference is so large.

The `depth` variable is supposedly computed via
`depth_computed = 100 * 2 * z / (x + y)`. Compute
`diff = depth - depth_computed`: This is a measure of discrepancy
between the given and computed depth. Additionally, compute the
*coefficient of variation* `cov = sd(x) / mean(x)` for both `depth` and
`diff`: This is a dimensionless measure of variation in a variable.
Assign the resulting tibble to `df_q4`, and assign the appropriate
values to `cov_depth` and `cov_diff`. Comment on the relative values of
`cov_depth` and `cov_diff`; why is `cov_diff` so large?

*Note*: Confusingly, the documentation for `diamonds` leaves out the
factor of `100` in the computation of `depth`.

```{r q4-task}
## TODO: Compute the `cov_depth` and `cov_diff` and assign the result to df_q4
df_q4 <-
  diamonds %>%
  mutate(
    depth_computed = 100 * 2 * z / (x + y),
    diff = depth - depth_computed
  ) %>%
  summarize(
    cov_depth = sd(depth, na.rm = TRUE) / mean(depth, na.rm = TRUE),
    cov_diff = sd(diff, na.rm = TRUE) / mean(diff, na.rm = TRUE)
  )
df_q4
```

**Observations**

-   Comment on the relative values of `cov_depth` and `cov_diff`.
    -   `cov_depth` is much smaller because the depths of the diamonds
        vary greatly with the sizes of diamonds throughout the data-set.
-   Why is `cov_diff` so large?
    -   `cov_diff` is much larger than `cov_depth` because the `diff` is
        a measure of error, rather than the actual size of the diamonds.
        Therefore, `diff` is much more closely varied, leading to a
        higher covariance.

The following test will verify that your `df_q4` is correct:

```{r q4-tests}
## NOTE: No need to change this!
assertthat::assert_that(abs(df_q4 %>% pull(cov_depth) - 0.02320057) < 1e-3)
assertthat::assert_that(abs(df_q4 %>% pull(cov_diff) - 497.5585) < 1e-3)
print("Nice!")
```

There is nonzero difference between `depth` and the computed `depth`;
this raises some questions about how `depth` was actually computed! It's
often a good idea to try to check derived quantities in your data, if
possible. These can sometimes uncover errors in the data!

### **q5** Compute and observe

Compute the `price_mean = mean(price)`, `price_sd = sd(price)`, and
`price_cov = price_sd / price_mean` for each `cut` of diamond. What
observations can you make about the various cuts? Do those observations
match your expectations?

```{r q5-task}
## TODO: Assign result to df_q5
df_q5 <-
  diamonds %>%
  group_by(cut) %>%
  summarize(
    price_mean = mean(price),
    price_sd = sd(price),
    price_cov = price_sd / price_mean
  )
df_q5
```

The following test will verify that your `df_q5` is correct:

```{r q5-tests}
## NOTE: No need to change this!
assertthat::assert_that(
  assertthat::are_equal(
    df_q5 %>%
      select(cut, price_cov) %>%
      mutate(price_cov = round(price_cov, digits = 3)),
    tibble(
      cut = c("Fair", "Good", "Very Good", "Premium", "Ideal"),
      price_cov = c(0.817, 0.937, 0.988, 0.949, 1.101)
    ) %>%
    mutate(cut = fct_inorder(cut, ordered = TRUE))
  )
)
print("Excellent!")
```

**Observations**:

-   I notice that the better the cut, the higher the price covariance.
-   Mean price is highest for Fair and Premium cuts, SD is highest for
    Premium, and Cov is highest for Ideal cut.

<!-- include-exit-ticket -->

# Exit Ticket

<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit
ticket survey**, [linked
here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-data02-derive-assignment.Rmd).
