---
title: "RMS Titanic"
author: "Brooke Moss"
date: 2023-01-27
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task
in analyzing a dataset is to understand trends as they vary across these
different variables. Unless we're careful and thorough, we can easily
miss these patterns. In this challenge you'll analyze a dataset with a
small number of categorical variables and try to find differences among
the groups.

*Reading*: (Optional) [Wikipedia
article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|------------------|-----------------------------|-------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight**
before the day of the class discussion of the challenge. See the
[Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true)
for more information.

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about
67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

-   Class \<chr\>
-   Sex \<chr\>
-   Age \<chr\>
-   Survived \<chr\>
-   n \<dbl\>

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

-   According to the Wikipedia page, there were 2224 souls aboard the
    Titanic, versus the 2201 observations in the data-set.
-   Are there any differences?
    -   Yes, the number of people reported on Wikipedia is 23 greater
        than in the data-set.
-   If yes, what might account for those differences?
    -   Unidentified people
    -   Last-minute cancellations
    -   Passengers with aliases got double-counted

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_titanic %>%
  filter(Survived == "Yes") %>%
  count(Class, Sex, wt = n) %>%
  ggplot() + geom_col(mapping = aes(x = Class, y = n, fill = Sex), position = "dodge")
```

**Observations**:

-   In my process, I noticed that I had to spend some time figuring out
    how to combine the "child" and "adult" rows for each category, so
    that the plot would be complete.
-   I also notice that the higher the class, the more surviving women,
    and the lower the class, the more surviving men (2nd class
    notwithstanding). I want to look at the data for all the souls on
    board to see if this is significant or just the breakdown of the
    passengers.

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for
understanding differences between groups. This is because the
differences we see in counts could be due to either the relative size of
the group OR differences in outcomes for those groups. To make
comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age`
group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot() + geom_col(mapping = aes(
    x = Class,
    y = Prop,
    fill = Sex
  ), position = "dodge")
```

**Observations**:

-   I notice that 100% of the 1st and 2nd class passengers in this plot
    have survived., along with most of the female crew.
-   Is there anything *fishy* going on in your plot?
    -   At first glance, I notice that the 1st-3rd class numbers do not
        line up with what I would expect, especially the 100% survival
        rate in 1st and 2nd class. This indicates that the plot is
        likely drawing both data-points on top of each other, with the
        bigger value masking the smaller value. This is because I did
        not combine the `n` values (and by extension `Prop` values) for
        the children and adults who otherwise would be the same
        category.

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider
additional variables!

```{r q5-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot() +
    geom_col(mapping = aes(
      x = Class,
      y = Prop,
      fill = Sex
    ), position = "dodge") +
    facet_grid(. ~ Age)
```

**Observations**:

-   When showing the two age groups side by side, I notice that 100% of
    the children, and the vast majority of women, in 1st and 2nd class
    survived, while a very small percentage of the men survived.
-   I notice, generally, that the higher the class, the more women
    survived.
-   If you saw something *fishy* in q4 above, use your new plot to
    explain the fishy-ness.
    -   In q4, I hadn't represented *or* removed the `Age` attribute.
        Back in q3, I removed the age attribute by adding the `n` values
        for the corresponding child and adult entries. In q5, I added
        `Age` as an aesthetic, by splitting children and adults into a
        `facet_grid`. This allows me to see the data for children and
        adults separately, and also compare them to each other.

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional
Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing
proportions is akin to non-dimensionalizing a quantity.
