---
title: "Aluminum Data"
author: "Brooke Moss"
date: 2023-02-12
output:
  github_document:
    toc: true
prerequisites:
  - e-data03-pivot-basics
---

*Purpose*: When designing structures such as bridges, boats, and planes,
the design team needs data about *material properties*. Often when we
engineers first learn about material properties through coursework, we
talk about abstract ideas and look up values in tables without ever
looking at the data that gave rise to published properties. In this
challenge you'll study an aluminum alloy dataset: Studying these data
will give you a better sense of the challenges underlying published
material values.

In this challenge, you will load a real dataset, wrangle it into tidy
form, and perform EDA to learn more about the data.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|------------------|-----------------------------|-------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight**
before the day of the class discussion of the challenge. See the
[Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true)
for more information.

```{r setup}
library(tidyverse)
```

*Background*: In 1946, scientists at the Bureau of Standards tested a
number of Aluminum plates to determine their
[elasticity](https://en.wikipedia.org/wiki/Elastic_modulus) and
[Poisson's ratio](https://en.wikipedia.org/wiki/Poisson%27s_ratio).
These are key quantities used in the design of structural members, such
as aircraft skin under [buckling
loads](https://en.wikipedia.org/wiki/Buckling). These scientists tested
plats of various thicknesses, and at different angles with respect to
the [rolling](https://en.wikipedia.org/wiki/Rolling_(metalworking))
direction.

# Loading and Wrangle

<!-- -------------------------------------------------- -->

The `readr` package in the Tidyverse contains functions to load data
form many sources. The `read_csv()` function will help us load the data
for this challenge.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/stang.csv"

## Load the data
df_stang <- read_csv(filename)
df_stang
```

Note that these data are not tidy! The data in this form are convenient
for reporting in a table, but are not ideal for analysis.

### **q1** Tidy `df_stang` to produce `df_stang_long`. You should have column names `thick, alloy, angle, E, nu`. Make sure the `angle` variable is of correct type. Filter out any invalid values.

*Hint*: You can reshape in one `pivot` using the `".value"` special
value for `names_to`.

```{r q1-task}
## TASK: Tidy `df_stang`
df_stang_long <-
  df_stang %>%
  pivot_longer(
    names_to = c(".value", "angle"),
    names_sep = "_",
    cols = starts_with("E") | starts_with("nu"),
    names_transform = list(angle = as.integer)
  ) %>%
  filter(E>0 | nu>0)

df_stang_long
```

Use the following tests to check your work.

```{r q1-tests}
## NOTE: No need to change this
## Names
assertthat::assert_that(
              setequal(
                df_stang_long %>% names,
                c("thick", "alloy", "angle", "E", "nu")
              )
            )
## Dimensions
assertthat::assert_that(all(dim(df_stang_long) == c(26, 5)))
## Type
assertthat::assert_that(
              (df_stang_long %>% pull(angle) %>% typeof()) == "integer"
            )

print("Very good!")
```

# EDA

<!-- -------------------------------------------------- -->

## Initial checks

<!-- ------------------------- -->

### **q2** Perform a basic EDA on the aluminum data *without visualization*. Use your analysis to answer the questions under *observations* below. In addition, add your own *specific* question that you'd like to answer about the data---you'll answer it below in q3.

```{r q2-task}
##
summary(df_stang_long)
group_by(df_stang_long, alloy) %>% summarize()
group_by(df_stang_long, angle) %>% summarize()
group_by(df_stang_long, thick) %>% summarize()
```

**Observations**:

-   Is there "one true value" for the material properties of Aluminum?
    -   There does not appear to be one "true value", as the properties
        seem to shift with the different samples and angles, judging by
        the result of running `summary(df_stang_long)`.
-   How many aluminum alloys are in this dataset? How do you know?
    -   By using `group_by(alloy) %>% summarize()`, I can see that there
        is only one group, meaning there is only one alloy in this
        dataset.
-   What angles were tested?
    -   When grouping by angle, I can see that there are 3 angles: 0,
        45, and 90.
-   What thicknesses were tested?
    -   By using the same `summarize` method from above, I can see that
        the tested thicknesses are 0.022, 0.032, 0.064, and 0.081.
-   *(Write your own question here)*
    -   (How) do the `E` and `nu` values correlate with the thickness of
        the aluminum?

## Visualize

<!-- ------------------------- -->

### **q3** Create a visualization to investigate your question from q2 above. Can you find an answer to your question using the dataset? Would you need additional information to answer your question?

```{r q3-task}
## TASK: Investigate your question from q2 here
df_stang_long %>%
  group_by(thick) %>%
  ggplot() +
  geom_boxplot(aes(
    x = as_factor(thick),
    y = E
  ))
df_stang_long %>%
  group_by(thick) %>%
  ggplot() +
  geom_boxplot(aes(
    x = as_factor(thick),
    y = nu
  ))
```

**Observations**:

-   It appears from my boxplots that E values generally are negatively
    correlated with thickness, but it does not appear to be an
    extraordinarily strong correlation.
-   For `nu`, the situation seems similar, although the nu values seem
    to gradually increase and then sharply drop off among the set of
    0.081 thickness samples.

### **q4** Consider the following statement:

"A material's property (or material property) is an intensive property
of some material, i.e. a physical property that does not depend on the
amount of the material."[2]

Note that the "amount of material" would vary with the thickness of a
tested plate. Does the following graph support or contradict the claim
that "elasticity `E` is an intensive material property." Why or why not?
Is this evidence *conclusive* one way or another? Why or why not?

```{r q4-vis}
## NOTE: No need to change; run this chunk
df_stang_long %>%

  ggplot(aes(nu, E, color = as_factor(thick))) +
  geom_point(size = 3) +
  theme_minimal()
```

**Observations**:

-   Does this graph support or contradict the claim above?
    -   I would not say this graph is conclusive one way or the other,
        but it does seem to suggest that elasticity is not an intensive
        material property, since at the very least, the E values for the
        0.081 thickness samples are distinctly lower than the rest.
        However, since the rest of the thicknesses are fairly jumbled as
        far as E values, I wouldn't say this is conclusive enough, as
        there could be some circumstances that cause 0.081 to be further
        away.

# References

<!-- -------------------------------------------------- -->

[1] Stang, Greenspan, and Newman, "Poisson's ratio of some structural
alloys for large strains" (1946) Journal of Research of the National
Bureau of Standards, (pdf
link)[<https://nvlpubs.nist.gov/nistpubs/jres/37/jresv37n4p211_A1b.pdf>]

[2] Wikipedia, *List of material properties*, accessed 2020-06-26,
(link)[<https://en.wikipedia.org/wiki/List_of_materials_properties>]
