---
title: "Gapminder"
author: "Brooke Moss"
date: 2023-02-22
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge
you'll further practice EDA by first completing a guided exploration,
then by conducting your own investigation. This challenge will also give
you a chance to use the wide variety of visual tools we've been
learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|-----------------|--------------------------|-----------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight**
before the day of the class discussion of the challenge. See the
[Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true)
for more information.

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is
an independent organization that seeks to educate people about the state
of the world. They seek to counteract the worldview constructed by a
hype-driven media cycle, and promote a "fact-based worldview" by
focusing on data. The dataset we'll study in this challenge is from
Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to
the high-level process we're going through---after this guided round
you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
gapminder
summary(gapminder)
```

**Observations**:

-   country
-   continent
-   year
-   lifeExp
-   pop
-   gdpPercap

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble.
(Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <-
  gapminder %>%
  pull(year) %>%
  max()
year_min <-
  gapminder %>%
  pull(year) %>%
  min()
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document
your challenges and attempt to produce the most informative visual you
can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent

# Final Plot
gapminder %>%
  filter(year == year_min) %>%
  ggplot() +
  geom_boxplot(mapping = aes(
    x = continent,
    y = gdpPercap,
  )) +
  scale_y_log10()

# Second Try
gapminder %>%
  filter(year == year_min) %>%
  ggplot() +
  geom_boxplot(mapping = aes(
    x = continent,
    y = gdpPercap,
  ))

# First Try
gapminder %>%
  filter(year == year_min) %>%
  ggplot() +
  geom_violin(mapping = aes(
    x = continent,
    y = gdpPercap,
  ))
```

**Observations**:

-   One data point within the Asian continent is astronomically higher
    than all of the other datapoints, I wonder which country that is.
-   Oceania has a very compressed set of points, which likely points to
    only a few countries being included.

**Difficulties & Approaches**:

-   Initial plot (violin) had such a long bar for Asia that the rest of
    the plot was unreadably tiny at the bottom.
-   Taking the boxplot instead of the violin gives a somewhat better
    view of the data, but the outlier still makes the rest of the plot
    difficult to understand.
-   By utilizing `scale_y_log10` to put the y axis on a log scale, the
    one extreme outlier no longer makes the rest of the plot unreadable.

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
gapminder %>%
  filter(year == year_min) %>%
  group_by(continent) %>%
  filter(continent != "Oceania") %>%
  filter(gdpPercap < quantile(gdpPercap, 0.01) | gdpPercap > quantile(gdpPercap, 0.99))
```

**Observations**:

-   Identify the outlier countries from q2
    -   Kuwait
    -   United States
    -   Switzerland

*Hint*: For the next task, it's helpful to know a ggplot trick we'll
learn in an upcoming exercise: You can use the `data` argument inside
any `geom_*` to modify the data that will be plotted *by that geom
only*. For instance, you can use this trick to filter a set of points to
label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple
variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>%
  filter(year %in% c(year_min, year_max)) %>%
  ggplot(aes(x = continent, y = gdpPercap)) +
  geom_boxplot() +
  facet_grid(rows = vars(year), scales = "free") +
  geom_point(
    data = . %>% filter(country %in% c("Kuwait", "United States", "Switzerland")),
    mapping = aes(color = country)
  )
```

**Observations**:

-   All 3 1952 outlier countries are still at or near the top of the GDP
    per capita for their respective continents.
-   However, Kuwait in particular more than halved their GDP per capita,
    leading them to now be much closer to the rest of the Asian
    distribution.
-   Switzerland is no longer the extreme outlier for Europe, rather
    being above the 75th percentile but not 99th.
-   The United States is still the highest in the Americas, and their
    GDP per capita has actually increased significantly, although they
    appear to be about at the same point relative to the Americas
    distribution as they were in 1952.

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP
per capita at two time points. You can continue looking at outliers,
consider different years, repeat the exercise with `lifeExp`, consider
the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph
gapminder %>%
  filter(country %in% c("Kuwait", "United States", "Switzerland")) %>%
  ggplot(aes(x = year, y = gdpPercap)) +
  geom_line(mapping = aes(color = country))
```

-   Although I already visualized the GDP per capita of the 3 outlier
    countries (USA, Kuwait, Switzerland) in both 1952 and 2007, I
    thought it would be useful to visualize their progression throughout
    the years.
-   From this plot, I can see that the USA and Switzerland seem to
    follow a very similar trajectory, perhaps indicative of a general
    trend in the international economy. After about 1980-1985, Kuwait
    also follows this trend.
-   The massive drop in Kuwait's economy between 1973 and 1985 coincides
    perfectly with the aftermath of the 1973 Oil Crisis, which tanked
    Kuwait's oil-dependent economy.

```{r q5-task2}
## TASK: Your second graph
gapminder %>%
  # filter(country %in% c("Kuwait", "United States", "Switzerland")) %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
  geom_point(mapping = aes(color = year))

gapminder %>%
  arrange(lifeExp)
```

-   Apart from being accidentally very pretty, this plot allows me to
    see a correlation between GDP per capita and life expectancy.
-   It is apparent that the highest life expectancy countries have
    relatively high GDP per capita. There are a few outlier countries
    with extremely high GDP that buck this trend, but these are likely
    autocracies (such as the Kuwaiti oil state) where wealth is
    concentrated in the leadership and therefore the population doesn't
    see life expectancy increases from the wealth.
-   In addition, I see that the coloration by year shows a stark
    increase in life expectancy overall as time goes on. Almost all of
    the high-life expectancy points are from recent years. However, as
    there is a large amount of overplotting on this figure, I can't
    reliably make this distinction without also creating a timeseries or
    other more focused plot.
-   I do notice one point at the very bottom left of the plot that is
    very recent and wonder what country it might be and whether its
    decline relates to some world events.
    -   After further investigation (sorting the table by lowest life
        expectancy), this lowest value seems to be from 1992 in Rwanda,
        which was during the Rwandan civil war (and just 2 years before
        what is commonly referred to as the Rwandan genocide).

```{r q5-task3}
## TASK: Your third graph
gapminder %>%
  group_by(continent, year) %>%
  summarize(meanLife = mean(lifeExp)) %>%
  ggplot(aes(x = year, y = meanLife)) +
  geom_line(aes(color = continent))
```

-   This plot gives a relatively simple representation of life
    expectancy in each continent over time. Each line is the mean of
    that continent's life expectancy over time.
-   I notice that between 1986 and 2002, the African life expectancy
    leveled off, while the other continents seemed to continue climbing.
