---
title: "SAT and College Grades"
author: "Brooke Moss"
date: 2023-04-28
output:
  github_document:
    toc: true
prerequisites:
  - e-vis00-basics
---

*Purpose*: How do we apply hypothesis testing to investigating data? In
this challenge you'll practice using hypothesis testing tools to make
sense of a dataset.

*Reading*: - [Harvard Study Says SATs Should Be Optional: Here's
Why](https://www.csmonitor.com/USA/USA-Update/2016/0120/Harvard-study-says-SATs-should-be-optional.-Here-s-why)
(Optional); easy-to-read news article on colleges going SAT-free -
[Norm-Referenced Tests and Race-Blind
Admissions](https://cshe.berkeley.edu/publications/norm-referenced-tests-and-race-blind-admissions-case-eliminating-sat-and-act-university)
(Optional); technical report on relationship between the SAT/ACT and
non-academic factors

*Credit*: This is based on a [case
study](http://onlinestatbook.com/2/case_studies/sat.html) originally
prepared by Emily Zitek, with data collected through the research of
Thomas MacFarland.

```{r setup}
library(tidyverse)
library(readxl)
library(broom)
library(modelr)
library(rsample)
library(correlation)
```

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|--------------|----------------------------|-------------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight**
before the day of the class discussion of the challenge. See the
[Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true)
for more information.

*Background*: Every year about 2 million students take the Scholastic
Aptitude Test (SAT). The exam is
[controversial](http://www.nea.org/home/73288.htm) but [extremely
consequential](https://www.csmonitor.com/2004/0518/p13s01-legn.html).
There are many claims about the SAT, but we're going to look at just
one: Is the SAT predictive of scholastic performance in college? It
turns out this is a fairly complicated question to assess---we'll get an
introduction to some of the complexities.

# Obtain the Data

<!-- -------------------------------------------------- -->

### **q1** Visit the [SAT and College GPA](http://onlinestatbook.com/2/case_studies/sat.html) case study page, scroll to the bottom, and click the `Open Data with Excel` button. This will allow you to download an `xls` file. Save the file to your `data` folder, load the data as `df_sat`, and perform your "first checks" against these data. Answer the questions below:

```{r q1-task}
## TODO:
df_sat <-
  read_xls(
    path = "./data/sat.xls"
  )

## TODO: Do your "first checks"
glimpse(df_sat)
summary(df_sat)
```

**Observations**:

-   Fill in the following "data dictionary"

| Column     | Meaning                                |
|------------|----------------------------------------|
| `high_GPA` | High school grade point average        |
| `math_SAT` | Math SAT score                         |
| `verb_SAT` | Verbal SAT score                       |
| `comp_GPA` | Computer science grade point average   |
| `univ_GPA` | Overall university grade point average |

-   What information do we have about these students?
    -   High school GPA, university GPA, computer science GPA, and SAT
        scores.
    -   All students were computer science students at a state
        university.
    -   All students graduated with a BS in CS.
-   What kinds of information *do we not have* about these students?
    -   Gender
    -   Race
    -   High school course options
    -   Family income
    -   High school funding/opportunities
    -   Hometown
-   Based on these missing variables, what possible effects could be
    present in the data that we would have *no way of detecting*?
    -   We have no way of detecting which students may have had
        additional resources to prepare for the SAT.
    -   We don't know which students may have come from school districts
        with limited access to advanced coursework.

# Analysis with Hypothesis Testing

<!-- ----------------------------------------------------------------------- -->

We're going to use two complementary approaches to analyze the data, the
first based on hypothesis testing of correlation coefficients, and the
second based on fitting a regression model and interpreting the
regression coefficients.

To simplify the analysis, let's look at a composite SAT score:

```{r compute-composite}
## NOTE: No need to edit this
df_composite <-
  df_sat %>%
  mutate(both_SAT = math_SAT + verb_SAT)
```

## View 1: Correlations

<!-- ----------------------------------------------------------------------- -->

### **q2** Create a *single* plot that shows `univ_GPA` against *both* `high_GPA` and `both_SAT`. Visually compare the two trends.

*Hint*: One way to do this is to first *pivot* `df_composite`.

```{r q2-task}
## TODO:
df_composite %>%
  pivot_longer(
    cols = c(high_GPA, both_SAT),
    names_to = "type",
    values_to = "value"
  ) %>%
  ggplot(aes(x = value, y = univ_GPA)) +
  geom_point() +
  facet_grid(cols = vars(type), scales = "free")
```

**Observations**:

-   What relationship do `univ_GPA` and `both_SAT` exhibit?
    -   `univ_GPA` and `both_SAT` exhibit a somewhat shallow linear
        relationship, with an outlier group with low of both types of
        scores. This group is pretty far away from the trend in the rest
        of the data.
    -   It seems that a high SAT predicts a high GPA, but a lower SAT
        doesn't necessarily predict much about GPA.
-   What relationship do `univ_GPA` and `high_GPA` exhibit?
    -   `univ_GPA` and `high_GPA` exhibit a much more linear
        relationship. Although there are some low-`high_GPA`,
        high-`univ_GPA` points, for the most part the `high_GPA` seems
        to predict the `univ_GPA`.

### Hypothesis Testing with a Correlation Coefficient

<!-- ------------------------- -->

We can use the idea of hypothesis testing with a correlation
coefficient. The idea is to set our null hypothesis to the case where
there is no correlation, and test to see if the data contradict that
perspective. Formally, the null (H0) and alternative (HA) hypotheses
relating to a correlation coefficient between two variables `X, Y` are:

$$\text{H0: } \text{Corr}[X, Y] = 0$$

$$\text{HA: } \text{Corr}[X, Y] \neq 0$$

The R function `cor.test` implements such a hypothesis test under the
assumption that `X, Y` are both normally distributed. First, let's check
to see if this assumption looks reasonable for our data.

### **q3** Plot histograms for `both_SAT, high_GPA, univ_GPA`. Which---if any---of the variables look approximately normally distributed.

```{r q3-task}
df_composite %>%
  pivot_longer(
    cols = c(high_GPA, both_SAT, univ_GPA),
    names_to = "type",
    values_to = "value"
  ) %>%
  ggplot() +
  geom_histogram(aes(x = value), bins = 15) +
  facet_grid(cols = vars(type), scales = "free")
```

**Observations**:

-   To what extent does `both_SAT` look like a normal distribution?
    -   `both_SAT` doesn't look very normal, although it looks close to
        being a skewed-right distribution.
-   To what extent does `high_GPA` look like a normal distribution?
    -   `high_GPA` looks to be the closest to a normal distribution,
        partially skewed-left.
-   To what extent does `univ_GPA` look like a normal distribution?
    -   `univ_GPA` looks like a normal distribution except for the
        additional spike between 2.0 and 2.5.

Keep in mind your findings as you complete q4.

### **q4** Use the function `cor.test()` to construct confidence intervals for `corr[high_GPA, univ_GPA]` and `corr[both_SAT, univ_GPA]`. Answer the questions below.

```{r q4-task}
## TODO: Use the function cor.test() to test the correlations between
##       high_GPA and univ_GPA, as well as between
##       both_SAT and univ_GPA
corr_GPA <-
  cor.test(
    x = df_composite$high_GPA,
    y = df_composite$univ_GPA
  )

corr_SAT <-
  cor.test(
    x = df_composite$both_SAT,
    y = df_composite$univ_GPA
  )

c(corr_GPA$estimate, corr_SAT$estimate)
c(corr_GPA$conf.int, corr_SAT$conf.int)
```

**Observations**:

-   Which correlations are significantly nonzero?
    -   Both correlation values are significantly nonzero.
-   Which of `high_GPA` and `both_SAT` seems to be more strongly
    correlated with `univ_GPA`?
    -   `high_GPA` appears to be more strongly correlated with
        `univ_GPA`, as its correlation value of 0.7795631 is greater
        than 0.6846776.
-   How do the results here compare with the visual you created in q2?
    -   These results appear to be consistent with my visual from q2, as
        `high_GPA` again has a higher correlation (presenting as a more
        linear relationship in q2).
-   Based on these results, what can we say about the predictive
    capabilities of both `high_GPA` and `both_SAT` to predict
    `univ_GPA`?
    -   From these results, we can say that `high_GPA` is a somewhat
        better predictor of `univ_GPA` than `both_SAT` is.

Finally, let's use the bootstrap to perform the same test using
*different* assumptions.

### **q5** Use the bootstrap to approximate a confidence interval.

Use the bootstrap to approximate a confidence interval for
`cor[high_GPA, univ_GPA]`. Compare your results---both the estimate and
confidence interval---to your results from q4.

*Hint 1*. The `cor(x, y)` function computes the correlation between two
variables `x` and `y`. You may find this more helpful than the
`cor.test()` function we used above.

*Hint 2*. You'll find that the documentation for `int_pctl` has some
**really** useful examples for this task!

```{r q5-task}
## TODO: Complete the following helper function to do a bootstrap analysis
corr_high_GPA <- function(split) {
  tibble(
    term = "cor",
    estimate = cor(df_composite$high_GPA, df_composite$univ_GPA)
  )
}
## Use the bootstrap to approximate a CI
df_composite %>%
  bootstraps(times = 1000) %>%
  mutate(estimates = map(splits, corr_high_GPA)) %>%
  int_pctl(estimates)
```

**Observations**:

-   How does your estimate from q5 compare with your estimate from q4?
    -   My estimate here is identical to the above estimate.
-   How does your CI from q5 compare with your CI from q4?
    -   In q4, my confidence interval was .15 wide, whereas in q5 it is
        nonexistent, with the lower and upper bounds both being
        identical to the estimate.

*Aside*: When you use two different approximations to compute the same
quantity and get similar results, that's an *encouraging sign*. Such an
outcome lends a bit more credibility to the results.

## View 2: Modeling

<!-- ------------------------- -->

Correlations are useful for relating two variables at a time. To study
the relationship among more variables we can instead use a fitted model.
Using a model, we can also help assess whether it is *worthwhile* to
measure a variable.

To begin, let's first split the data into training and validation sets.

```{r split}
## NOTE: No need to edit
set.seed(101)

df_train <-
  df_composite %>%
  rowid_to_column() %>%
  slice_sample(n = 80)

df_validate <-
  df_composite %>%
  rowid_to_column() %>%
  anti_join(
    .,
    df_train,
    by = "rowid"
  )
```

### Hypothesis Testing with a Model

<!-- ------------------------- -->

We can combine the ideas of hypothesis testing with a model. Using a
model, we can express our hypotheses in terms of the model parameters.
For instance, if we were interested in whether $X$ has an affect on $Y$,
we might set up a model:

$$Y_i = \beta X_i + \epsilon_i$$

With the hypotheses:

$$\text{H0}: \beta = 0$$

$$\text{HA}: \beta \neq 0$$

In this case, we're testing for whether $X$ has a significant effect on
$Y$. Let's apply this idea to relating the variables `univ_GPA` and
`high_GPA`. Luckily R has built-in tools to construct a confidence
interval on the $\beta$'s in a regression [1]; we'll simply use those
tools rather than do it by hand.

### **q6** Fit a linear model predicting `univ_GPA` with the predictor `both_SAT`. Assess the model to determine how effective a predictor `both_SAT` is for `univ_GPA`. Interpret the resulting confidence interval for the coefficient on `both_SAT`.

```{r q6-task}
## TODO: Fit a model of univ_GPA on the predictor both_SAT
fit_basic <-
  df_composite %>% 
  lm(
    data = .,
    formula = univ_GPA ~ both_SAT
  )

rsquare(fit_basic, df_composite)

## NOTE: The following computes confidence intervals on regression coefficients
fit_basic %>%
  tidy(
    conf.int = TRUE,
    conf.level = 0.99
  )
```

**Observations**:

-   What is the confidence interval on the coefficient of `both_SAT`? Is
    this coefficient significantly different from zero?
    -   The confidence interval is [0.00198, 0.00349], with an estimate
        of 0.00274. This is extremely close to zero, but since zero is
        not within the interval, this is statistically significant.
-   By itself, how well does `both_SAT` predict `univ_GPA`?
    -   By itself, `both_SAT` is not a good predictor of `univ_GPA`.
        With an rsquared value of 0.468, it doesn't appear to be
        completely uncorrelated, but it is not well enough correlated to
        be a good predictor.

Remember from `e-model03-interp-warnings` that there are challenges with
interpreting regression coefficients! Let's investigate that idea
further.

### **q7** Fit a model predicting `univ_GPA` using both `high_GPA` and `both_SAT`. Compare the prediction accuracy and hypothesis test results.

```{r q7-task}
## TODO: Fit and assess models with predictors both_SAT + high_GPA, and high_GPA alone
fit_high <-
  df_composite %>% 
  lm(
    data = .,
    formula = univ_GPA ~ high_GPA
  )

fit_high %>%
  tidy(
    conf.int = TRUE,
    conf.level = 0.99
  )

fit_both <-
  df_composite %>% 
  lm(
    data = .,
    formula = univ_GPA ~ both_SAT + high_GPA
  )

fit_both %>%
  tidy(
    conf.int = TRUE,
    conf.level = 0.99
  )

rsquare(fit_high, df_composite)
rsquare(fit_both, df_composite)
```

**Observations**:

-   How well do these models perform, compared to the one you built in
    q6?
    -   The model with just `high_GPA` seems to perform better than the
        `both_SAT` model, with an rsquare value of 0.608, as compared to
        the 0.468 of `both_SAT`. When combining both `both_SAT` and
        `high_GPA` into one model, the rsquare improves slightly more to
        0.623, but judging by the coefficients, `high_GPA` has a much
        greater impact in that model than `both_SAT`.
-   What is the confidence interval on the coefficient of `both_SAT`
    when including `high_GPA` as a predictor?? Is this coefficient
    significantly different from zero?
    -   The confidence interval on the coefficient of `both_SAT` is
        [-.000223, 0.00181], which is not significantly different from
        zero.
-   How do the hypothesis test results compare with the results in q6?
    -   These results demonstrate greater prediction ability using
        `high_GPA` or a combination of `high_GPA` and `both_SAT` as
        compared to `both_SAT`. `high_GPA` is much more directly
        correlated with `univ_GPA`.

## Synthesize

<!-- ------------------------- -->

Before closing, let's synthesize a bit from the analyses above.

### **q8** Using the results from all previous q's, answer the following questions.

**Observations**:

-   Between `both_SAT` and `high_GPA`, which single variable would you
    choose to predict `univ_GPA`? Why?
    -   If I had to choose a single variable, I would likely choose
        `high_GPA`, because the plots and early analysis in this
        assignment seem to suggest it is more strongly correlated,
        especially the correlation calculations in q4. The model
        analysis in q6 and q7 also show that `high_GPA` is a better
        predictor, although it is not perfect.
-   Is `both_SAT` an effective predictor of `univ_GPA`? What specific
    pieces of evidence do you have in favor of `both_SAT` being
    effective? What specific pieces of evidence do you have against?
    -   According to q6 and q7, `both_SAT` appears to actually be an
        okay predictor of `univ_GPA`, though the earlier questions,
        especially the plots, seem to suggest otherwise. In q7
        especially, we can see that `both_SAT` can be used in
        *combination* with `high_GPA` to improve the model slightly.

# End Notes

<!-- ----------------------------------------------------------------------- -->

[1] There are also assumptions underlying this kind of testing, for more
information see this [Wiki
article](https://en.wikipedia.org/wiki/Linear_regression#Assumptions).
